---
title: "Case Study 1"
author: "Trace Smith"
date: "October 20, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction
 

### Downloading the Dataset:

There are two options when downloading the either of the datasets. First option is shown below where we identify the URL link to the data source and download the file from the website. Furthermore, `destfile` refers to directory the file will be stored in along with the name of the file (i.e. GDP.csv). The important piece to rememebr here is setting the working directorty to the root directory: `Case Study 1`.

```{r}
url<-"https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"
download.file(url,destfile="Data/FEDSTATS_Country.raw.csv")
```

```{r}
#Download Educational Data
url<-"https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"
download.file(url,destfile="Data/FGDP.raw.csv")
```

The second option is to download the file from [`The World Bank`](http://www.worldbank.org/) website directly and then move to from the download fold to the Data directory.

### Tyding the Dataset:



### Merging the Dataset:



### Question 1:
**Merge the data based on the country shortcode. How many of the IDs match?**




### Question 2:
**Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?**



### Question 3:
**What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?**



### Question 4:
**Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.**



### Question 5:
**Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group. How many countries are Lower middle income but among the 38 nations with highest GDP?**



### Conclusion